---
title: "Structural controllability of regime shifts"
author: Juan Rocha
output:
  html_notebook:
    toc: yes
    toc_float: yes
    highlight: tango
    code_folding: hide
    df_print: paged
    theme: 
      bootswatch: cosmo
      code_font:
        google: Fira Code
---

```{r setup, include = FALSE}
set.seed(12345)
library(tidyverse)
library(igraph)
library(reticulate)

use_python(python = "/anaconda3/bin/python")
#use_condaenv()
py_config()
# py_install("igraph")
# import("igraph")

# knitr::opts_chunk$set(
#   cache = TRUE, warning = FALSE, message = FALSE, echo = TRUE,
#   fig.width = 8, fig.height = 5
# )

```

Fist I read the data and keep only the regime shift 'coral transitions' as an
example.

```{r message=FALSE, warning=FALSE}
### Read CLD data# load CLD
dat <- read.csv2('~/Documents/JUAN/PhD-SRC/RS20+Analysis/RS_CLD_2018.csv') %>% 
  janitor::clean_names() %>% as_tibble()
dat <- dat[,-1]
dat$col <- ifelse(dat$polarity == -1, "#FC8D62","#1F78B4")

g <- filter(dat, regime_shift == "Coral transitions") %>% 
    droplevels() %>% 
    select(tail:regime_shift)
G <- graph_from_data_frame(d = g, directed = TRUE)

# as_edgelist(G)
# vertex_names <- unique(c(as.character(g$tail), as.character(g$head)))
 
g
```

Now `dat` is the dataset, the object is a `r class(dat)`. `G` is the network
object in igraph. I have written a function in python to calculate the feedback
arc set based on it's igraph functionality `fas.py`.

```{python}
import igraph
import pandas as pd

def feedback_set(dat):
    edgelist = list(dat[["tail", "head"]].itertuples(index=False, name = None))
    net = Graph.TupleList(edges = edgelist, directed = True)
    fas = Graph.feedback_arc_set(net)
    return fas
```

## Feedback vertex sets:

Testing on the coral reefs example, the feedback vertex set is:

```{r}
source_python("fas.py")
g.py <- r_to_py(g)
FAS <- feedback_set(dat = g.py)
# Here I need fas+1 because of python indexing, and the 1rst column because it's the source of the link. According to Zhou 2016 FVS and FAS problems are equivalent. My interpretation is that solving the FAS problem gives the origin vertex of the FVS problem.
as_edgelist(G)[FAS+1, 1]
```

Remember than indexing in python starts in zero while in R it stars in 1. That's
why to recover the list of edges that belong to the feedback arc set (FAS) one
needs to add 1 to the object in R.

Now that I have a working example, let's try to scale it up.

```{r warning=FALSE}
fas_all <- dat %>%
    split(.$regime_shift) %>%
    map(r_to_py) %>%
    map(feedback_set)

## correct the indexing from python to R by adding 1
fas_all <- lapply(fas_all, function(x) x + 1)

net_all <- dat %>%
    split(.$regime_shift) %>%
## Note: I need to use the ~ to tell map to do the select in each dataframe. Otherwise I get an error
    map(~select(., 1:4)) %>%
    map(graph_from_data_frame, directed = TRUE)


##  I don't need vetex names remember? the Feedback Arc Set gives me the arcs (edges), 
## then I should extract the unique source nodes.
edges_all <- net_all %>%
    map(as_edgelist) %>%
    map(as_tibble) %>%
    map(~pull(., V1))

vertices_control <- map2(edges_all, fas_all, `[`) %>%
    map(unique)

```

### Testing

Do a test to see if it works: Delete the vertices found and then check if the
remaining network is a tree or a directed acyclic graph (DAG).

```{r}
## delete vertices for one graph

g_test <- delete_vertices(
    graph = net_all[[1]],
    v = which(V(net_all[[1]]) %>% names() %in% vertices_control[[1]])
)

## test if it's a tree or a DAG
g_test %>% is.dag()
```

Now for all graphs:

```{r warning=FALSE}
test_all <-map2(.x = net_all, .y = vertices_control, function(x,y){
        delete_vertices(
            graph = x,
            v = which(names(V(x)) %in% y)
        )
    })

test_all %>% map(., function(x)is.dag(x)) %>% all()
```

### Visualizations

```{r}
df_fvs <- tibble(
    regime_shift = names(vertices_control),
    fvs = map(vertices_control, function(x) x) 
) %>% 
    unnest(fvs)

```

```{r}
df_fvs %>%
    unique() %>% 
    group_by(regime_shift) %>% 
    summarise(n_fvs = n()) %>% 
    mutate(regime_shift = as_factor(regime_shift) %>% 
               fct_reorder(., n_fvs, sort)) %>% 
    ggplot(aes(regime_shift, n_fvs)) +
    geom_col() +
    coord_flip() +
    labs(x = "Regime shifts", y = "Feedback vertex set size") +
    theme_light(base_size = 10)
```

```{r}
df_fvs %>% 
    group_by(fvs) %>% 
    count() %>% 
    arrange(desc(n)) 
```

I think one needs the unmatched nodes as well.

## Matching

First convert all networks to bipartite by using the edgelists of the CLDs.
Instead of deleting vertices, delete the edges of the feedback arc set first.

```{r}
# Delete the edges from the feedback arc set:

dags_all <- map2(net_all, fas_all, delete_edges)

dags_all %>% map(., function(x)is.dag(x)) %>% all()
```

```{r}
bip_all <- dat %>%
    mutate(
        tail = str_c(tail, "t", sep = "_"),
        head = str_c(head, "h", sep = "_")
    ) %>% 
    split(.$regime_shift) %>%
## Note: I need to use the ~ to tell map to do the select in each dataframe. Otherwise I get an error
    map(~select(., 1:4)) %>%
    map(graph_from_data_frame, directed = TRUE) %>% 
    # delete the links of the feedback arc set: Although names are different, the order of links is the same
    map2(fas_all, delete_edges) %>% 
    map(., function(x) {
        V(x)$type <- V(x) %>%
            names() %>% str_detect("_t")
        return(x)
    })

```

```{r}
matching <-  bip_all %>% 
    map(max_bipartite_match)

# minimum driver node set:
df_mds <- tibble(
    regime_shift = names(vertices_control),
    mds = map(matching, function(x) names(x$matching[is.na(x$matching)]) ) 
) %>% 
    unnest(mds) %>% 
    mutate(mds = str_remove_all(mds, pattern = "_t|_h"))
```

```{r}
df_mds %>%
    unique() %>% 
    group_by(regime_shift) %>% 
    summarise(n_mds = n()) %>% 
    mutate(regime_shift = as_factor(regime_shift) %>% 
               fct_reorder(., n_mds, sort)) %>% 
    ggplot(aes(regime_shift, n_mds)) +
    geom_col() +
    coord_flip() +
    labs(x = "Regime shifts", y = "Minimum driver nodes") +
    theme_light(base_size = 10)
```

## Minimum driving nodes and feedback vertex set

Combine the feedback vertex set and the minimum driver node set:

```{r}
df_fvs <- df_fvs %>% 
    rename(node = fvs) %>% 
    mutate(type = "fvs")

df_mds <- df_mds %>% 
    rename(node = mds) %>% 
    mutate(type = "mds")

df_comb <- bind_rows(
    unique(df_fvs), 
    unique(df_mds))

df_comb %>%
    group_by(regime_shift, node) %>% 
    add_count( ) %>% 
    arrange(desc(n))
```

Notice that some nodes are part of the feedback vertex set and the minimum
driver set once the network is made a DAG. To avoid double counting, I delete
them from being counted as MDS since they are already part of the FVS.

```{r}
df_comb <- df_comb %>% 
    group_by(regime_shift, node) %>% 
    add_count( ) %>% 
    filter((type != "mds" | n != 2)) 
```

```{r}
df_comb %>% 
    ungroup() %>% 
    group_by(regime_shift, type) %>% 
    summarize(n = n()) %>% 
    ungroup() %>% group_by(regime_shift) %>% 
    mutate(total = sum(n)) %>% 
    ggplot(aes(regime_shift, n)) +
    geom_col(aes(fill = type), position = "stack") +
    coord_flip() +
    scico::scale_fill_scico_d(palette = "berlin") +
    theme_linedraw()
```

```{r}
net_all[[1]] %>% V() %>% length()
```

Next steps:

1.  Add graph statistics: number of vertices and then proportion of them than
    are part of the minimum driver set.
2. Idea for figure 1 of the paper: a) number of driver set, b) proportion of driver set per RS, c) key variables across RS. Make a draft here but save the necessary datasets from the analysis and keep a Figures script aside.

```{r}
df_stats <- tibble(
  regime_shift = names(vertices_control),
  nodes = map_dbl(net_all, function(x) length(get.vertex.attribute(x)$name))
)

base::save(
  df_stats, df_fvs, df_mds, df_comb,
  file = "data_processed/individualRS_control.RData"
)
```

